Loop Invariant Code Motion in PyPy's Tracing JIT

### 2011-06-12 talk/iwtc11/paper.tex
various fixes/rewrites

\begin{figure}
\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)
jump($l_0$, $p_{0}$, $p_{5}$)
\end{lstlisting}
\caption{An Unoptimized Trace of the Example Interpreter}
\label{fig:unopt-trace}
\end{figure}

@@

\subsection{Loop peeling}
Loop peeling is achieved by inlining the trace at the end of
itself. The input arguments of the second iteration are replaced with
the jump arguments of the first iterations and then the arguments of all
the operations are updated to operate on the new input arguments. To
keep the single-assignment form new variables has to be introduced as
the results of all the operations. The first iteration of the loop
will end with a jump to the second iteration of the loop while the
second iteration will end with a jump to itself. This way the first
copy of the trace only be executed once while the second copy will be
used for every other iteration. The rationality here is that the
optimizations below typically will be able to optimize the second copy
more efficiently than the first. The trace from Figure~\ref{fig:unopt-trace} would
after this operation become the trace in Figure~\ref{fig:peeled-trace}.

@@

When applying the following optimizations to this two iteration trace
some care has to taken as to how the jump arguments of both
iterations and the input arguments of the second iteration are
treated. It has to be ensured that the second iteration stays a proper
trace in the sens that the operations within it only operations on
variables that are either among the input arguments of the second iterations
or are produced within the second iterations. To ensure this we need
to introduce a bit of formalism.

The original trace (prior too peeling) consists of three parts. 
A vector of input
variables, $I=\left(I_1, I_2, \cdots, I_{|I|}\right)$, a list of non
jump operations and a single
jump operation. The jump operation contains a vector of jump variables,
$J=\left(J_1, J_2, \cdots, J_{|J|}\right)$, that are passed as the input variables of the target loop. After
loop peeling there will be a second copy of this trace with input
variables equal to the jump arguments of the first copy, $J$, and jump
arguments $K$. Looking back at our example we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      I &=& \left( p_0, p_1 \right) \\
      J &=& \left( p_0, p_5 \right) \\
      K &=& \left( p_0, p_9 \right) \\
    \end{array}
  %\right.
  .
\end{equation}
To construct the second iteration from the first we also need a
function, $m$, mapping the variables of the first iteration onto the
variables of the second. This function is constructed during the
inlining. It is initialized by mapping the input arguments, $I$, to
the jump arguments $J$,
\begin{equation}
  m\left(I_i\right) = J_i \ \text{for}\ i = 1, 2, \cdots |I| .
\end{equation}
In the example that means (XXX which notation do we prefer?)
\begin{equation}
  m(v) = 
  \left\{
    \begin{array}{lcl}
      p_0 &\text{if}& v=p_0 \\
      p_5 &\text{if}& v=p_1 \\
    \end{array}
  \right.
  .
\end{equation}
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      m\left(p_0\right) &=& p_0 \\
      m\left(p_1\right) &=& p_5
    \end{array}
  %\right.
  .
\end{equation}
Each operation in the trace is inlined in the order they are
executed. To inline an operation with argument vector 
$A=\left(A_1, A_2, \cdots, A_{|A|}\right)$ producing the variable $v$
a new variable, $\hat v$ is introduced. The inlined operation will
produce $\hat v$ from the input arguments 
\begin{equation}
  \left(m\left(A_1\right), m\left(A_2\right), 
    \cdots, m\left(A_{|A|}\right)\right) . 
\end{equation}
Before the
next operation is inlined, $m$ is extend by making $m\left(v\right) = \hat
v$. After all the operations in the example have been inlined we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      m\left(p_0\right) &=& p_0 \\
      m\left(p_1\right) &=& p_5 \\
      m\left(i_2\right) &=& i_6 \\
      m\left(i_3\right) &=& i_7 \\
      m\left(i_4\right) &=& i_8 \\
      m\left(p_5\right) &=& p_9 \\
    \end{array}
  %\right.
  .
\end{equation}

\subsection{Redundant guard removal}
No special concerns needs to be taken when implementing redundant
guard removal together with loop peeling. However the the guards from
the first iteration might make the guards of the second iterations
redundant and thus removed. So the net effect of combining redundant
guard removal with loop peeling is that guards are moved out of the
loop. The second iteraton of the example reduces to

@@

\subsection{Heap caching}
The objective of heap caching is to remove \lstinline{get} and
\lstinline{set} operations whose results can be deduced from previous
\lstinline{get} and \lstinline{set} operations. Exact details of the
process are outside the scope of this paper We will here assume that
it works perfectly and only consider the interactions with loop
peeling.

@@

After that, the second
iteration will no longer be proper as it operates on $i_3$ and $i_4$
which are not part of it. The solution is to extend the input
arguments, $J$, with those two variables. This will also extend the
jump arguments of the first iteration, which is also $J$. 
Implicitly that also extends the jump arguments of the second iteration, $K$,
since they are the inlined versions of $J$. That is the, $I$ has to
be replaced by $\hat I$ which is formed as a concatenation of $I$ and
$\left(i_3, i_4\right)$. At the same time $K$ has to be replaced by
$\hat K$ which is formed as a concatenation of $K$ and 
$\left(m\left(i_3\right), m\left(i_4\right)\right) = \left(i_7, i_8\right)$. 
The variable $i_7$ will then be replaced by $i_3$ by the heap caching
algorithm as it has removed the variable $i_7$. XXX: Maybe we should
replace $i_7=$get(...) with $i_7=i_3$ instead of removing it?

In general what is needed is for the heap optimizer is to keep track of
which variables from the first iterations it reuses in the second
iteration. It has to construct a vector of such variables $H$ which
can be used to update the input and jump arguments,
\begin{equation}
  \hat J = \left(J_1, J_2, \cdots, J_{|J|}, H_1, H_2, \cdots, H_{|H}\right)
\end{equation}
\begin{equation}
  \hat K = \left(K_1, K_2, \cdots, K_{|J|}, m(H_1), m(H_2), \cdots, m(H_{|H})\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into

@@

\subsection{Virtualization}
Using escape analysis we can XXX

Let $\tilde J$ be all variables in $J$ not representing virtuals (in the
same order). Extend it with all non virtual fields, $H_i$, of the
removed virtuals,
\begin{equation}
  \hat J = \left(\tilde J_1, \tilde J_2, \cdots, \tilde J_{|\tilde J|}, 
                 H_1, H_2, \cdots, H_{|H}\right)
\end{equation}
and let
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}

>>>>

\begin{figure}
XXX the code for is\_positive is missing everywhere
\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)
jump($l_0$, $p_{0}$, $p_{5}$)
\end{lstlisting}
\caption{An Unoptimized Trace of the Example Interpreter}
\label{fig:unopt-trace}
\end{figure}

@@

\subsection{Loop Peeling}

XXX find reference

Loop peeling is achieved prefixing the loop with one iteration of itself. The
peeled of iteration of the loop will end with a jump to the full loop, which
ends with a jump to itself. This way the peeled of iteration will only be
executed once while the second copy will be used for every further iteration.
The trace from Figure~\ref{fig:unopt-trace} would after this operation become
the trace in Figure~\ref{fig:peeled-trace}.

@@

When applying the following optimizations to this two-iteration trace
some care has to taken as to how the jump arguments of both
iterations and the input arguments of the second iteration are
treated. It has to be ensured that the second iteration stays a proper
trace in the sense that the operations within it only operations on
variables that are either among the input arguments of the second iterations
or are produced within the second iterations. To ensure this we need
to introduce a bit of formalism.

The original trace (prior too peeling) consists of three parts. 
A vector of input
variables, $I=\left(I_1, I_2, \cdots, I_{|I|}\right)$, a list of non-
jump operations and a single
jump operation. The jump operation contains a vector of jump variables,
$J=\left(J_1, J_2, \cdots, J_{|J|}\right)$, that are passed as the input variables of the target loop. After
loop peeling there will be a second copy of this trace with input
variables equal to the jump arguments of the peeled copy, $J$, and jump
arguments $K$. Looking back at our example we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      I &=& \left( p_0, p_1 \right) \\
      J &=& \left( p_0, p_5 \right) \\
      K &=& \left( p_0, p_9 \right) \\
    \end{array}
  %\right.
  .
\end{equation}
To construct the second iteration from the first we also need a
function $m$, mapping the variables of the first iteration onto the
variables of the second. This function is constructed during the
inlining. It is initialized by mapping the input arguments, $I$, to
the jump arguments $J$,
\begin{equation}
  m\left(I_i\right) = J_i \ \text{for}\ i = 1, 2, \cdots |I| .
\end{equation}
In the example that means:

\begin{equation}
  %\left\{
    \begin{array}{lcl}
      m\left(p_0\right) &=& p_0 \\
      m\left(p_1\right) &=& p_5
    \end{array}
  %\right.
  .
\end{equation}

Each operation in the trace is inlined in order.
To inline an operation $v=op\left(A_1, A_2, \cdots, A_{|A|}\right)$
a new variable, $\hat v$ is introduced. The inlined operation will
produce $\hat v$ from the input arguments 
\begin{equation}
  \hat v = op\left(m\left(A_1\right), m\left(A_2\right), 
    \cdots, m\left(A_{|A|}\right)\right) . 
\end{equation}
Before the
next operation is inlined, $m$ is extend by assigning $m\left(v\right) = \hat
v$. For the example above, after all the operations have been inlined we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      m\left(p_0\right) &=& p_0 \\
      m\left(p_1\right) &=& p_5 \\
      m\left(i_2\right) &=& i_6 \\
      m\left(i_3\right) &=& i_7 \\
      m\left(i_4\right) &=& i_8 \\
      m\left(p_5\right) &=& p_9 \\
    \end{array}
  %\right.
  .
\end{equation}

\subsection{Redundant Guard Removal}

No special concerns needs to be taken when implementing redundant
guard removal together with loop peeling. The guards from
the first iteration might make the guards of the second iterations
redundant and thus removed. Therefore the net effect of combining redundant
guard removal with loop peeling is that loop-invariant guards are moved out of the
loop. The second iteration of the example reduces to

@@

\subsection{Heap Caching}
The objective of heap caching is to remove \lstinline{get} and
\lstinline{set} operations whose results can be deduced from previous
\lstinline{get} and \lstinline{set} operations. Exact details of the
process are outside the scope of this paper. We only consider the interaction
with loop peeling.

@@

After that, the second
iteration will no longer be in SSA form as it operates on $i_3$ and $i_4$
which are not part of it. The solution is to extend the input
arguments, $J$, with those two variables. This will also extend the
jump arguments of the first iteration, which is also $J$. 
Implicitly that also extends the jump arguments of the second iteration, $K$,
since they are the inlined versions of $J$. For the example $I$ has to
be replaced by $\hat I$ which is formed as a concatenation of $I$ and
$\left(i_3, i_4\right)$. At the same time $K$ has to be replaced by
$\hat K$ which is formed as a concatenation of $K$ and 
$\left(m\left(i_3\right), m\left(i_4\right)\right) = \left(i_7, i_8\right)$. 
The variable $i_7$ will then be replaced by $i_3$ by the heap caching
optimization as it has removed the variable $i_7$. XXX: Maybe we should
replace $i_7=$get(...) with $i_7=i_3$ instead of removing it?

In general what is needed is for the heap optimizer is to keep track of
which variables from the first iterations it reuses in the second
iteration. It has to construct a vector of such variables $H$ which
can be used to update the input and jump arguments,
\begin{equation}
  \hat J = \left(J_1, J_2, \cdots, J_{|J|}, H_1, H_2, \cdots, H_{|H}\right)
\end{equation}
\begin{equation}
  \hat K = \left(K_1, K_2, \cdots, K_{|J|}, m(H_1), m(H_2), \cdots, m(H_{|H})\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized to:

@@

\subsection{Allocation Removals}
Using escape analysis we can XXX

Let $\tilde J$ be all variables in $J$ not representing virtuals (in the
same order). Extend it with all non-virtual fields, $H_i$, of the
removed virtuals,
\begin{equation}
  \hat J = \left(\tilde J_1, \tilde J_2, \cdots, \tilde J_{|\tilde J|}, 
                 H_1, H_2, \cdots, H_{|H}\right)
\end{equation}
and let
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
